### 2021-08-06 reviewer_answers/Response to reviewers.tex

\item
These were my comments. I appreciate the paper, I enjoyed reading.

AUTHOR answer: Thank you for the positive evaluation of our manuscrupt, we really appreciated your thoughtful comments!
\end{enumerate}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{Referee \#2  (Remarks to the Author):}\\

This manuscript is concerned with a mathematical modelling for speciation using a general Lotka-Volterra equation modelling the evolution of some ecosystem. This model takes into account species traits and mutations through a diffusion term. This general equation is coupled with a system of ODE modelling the species level: density of species together with the evolution of the mean traits and the covariance matrix.

\begin{enumerate}
\item
The authors discuss and propose a model for the speciation processes leading to new species. To do so, they make use of a discretized version of the above equations. The discussions on the speciation processes seems to be highly related to the choice of the discrete system and in particular on the reconstruction operator. No connection with the original continuous system is clearly explained neither how it depends on the discretization.
Speciation is modeled in term of clustering in the trait space. Does it depend on the traits kept in the model?

@@

AUTHOR answer:

\item
From a modelling view point, how do you choose the selection function $\alpha=\alpha(x,y,t)$? How is it related to the coefficients $A_{i,j}$ in (2.5a)? How the multi-scale parameters are coupled?

AUTHOR answer: $A_{i,j}$ is related to $\alpha$ through eq. (2.6c), i.e., the effect of species $i$ on species $j$ (as opposed to how individuals with different trait coordinates effect eachother, as is the case at the population level). \\

The strategy for coupling the species and population scales in the multi-scale framework is outlined in section 5 "Speciation".
\end{enumerate}

### 2021-08-09 reviewer_answers/Response to reviewers.tex

\textbf{ Referee \#1  (Remarks to the Author)}\\

The paper deals with the process of speciation, which is an inportant process within evolutionary biology. The emphasis of the paper is on the posterior error estimation in a multi-scale framework. The paper is well-written and well-organised. The mathematical theories that have been developed are sound. I do not necessarily agree with the modelling framework, but this is not what the current manuscript is about, hence this remark of mine is irrelevant. I recommend minor revisions, and to aid the Authors present their work more clearly, I have the following points that are to be addressed:

\begin{enumerate}
\item
In general, the abstract may put more emphasis on what the paper is actually about: namely the posterior error analysis. This has only been written in the very last sentences of the abstract. Probably, this is a matter of taste ... \\

AUTHOR answer: We agree with this point. The entire abstract has been re-written to better reflect the actual contents of the paper.

@@

\item
Line 528 please, replace 'The figure 2 below' with 'Figure 2 below'

AUTHOR answer: This has been corrected.

\item
Your analysis is also built on the existence of statistical moments, hence in the case of a Cauchy distribution (which has a very 'normal' appearance), the model may lead to problems. It could be interesting to do a numerical experiment with a Cauchy distribution function. ($f(x) = 1/\pi 1 / (x^2 + 1)$ is the standard Cauchy distribution, which has no moments (average, standard deviation ...).

AUTHOR answer: The derivation of the species level model from the population level model rests on two fundamental assumptions; (1) that distinct species can be identified at all times over the global population distribution (i.e., as compactly supported, non-overlapping sub-distributions), and (2) that these sub-distributions are of a known statistical quality (in this case, the Gaussian/normal distribution). As discussed in [1] (where the species level model is derived), the choice of the Gaussian was to a certain degree due to convenience (although individual traits are commonly assumed to follow a normal distribution in the biological literature), hence other statistical distributions could also be accomodated (at the cost of an increasingly complex calculation/notation, see e.g., page 5-6 in [1] where the possibility of using the uniform distribution is briefly discussed). However, for the exact reasons given in your comment, the Cauchy distribution is not a possible choice, although Cauchy distributions do not pose a problem in terms of the population level model. \\

In the manuscript we emphasized the fact that speciation events will invalidate the species level model (violation of assumption (1)), but it may also be invalidated by the species distributions being of an unknown quality (or, with non-existing moments, e.g., Cauchy), hence violating assumption (2). It appears we have neglected to mention this. In the fifth paragraph of the introduction (pages 2-3), where we discuss the species and population level models, we have added a few sentences outlining the above point concerning assumption (2). We have also added in section (2.3) that here $n$ should be smooth enough for these moments to exist.\\

%These assumptions may be quite restrictive when modeling an eco-evolutionary system, but such is the cost when averaging the global trait-dependent population distribution/density into a state vector (perhaps coupling the population and species levels in a multi-scale framework might be worth a try...).  \\

@@

\item
line 664: The Authors note that, indeed, the Gaussian does not satisfy the boundary conditions. Indeed, if the boundaries are far enough, then, the boundary conditions are approximately satisfied. Hence, then there is no problem. The Authors could use the classical bound for a normal distribution (which is much sharper than Chebychev's Inequality) that

$$P(|X| \ge t ) \le \sqrt{\frac{2}{\pi}} \frac{\sigma}{t} exp(-\frac{t^2}{2 \sigma^2}),$$

where $X \sim \mathcal{N}(0,\sigma^2)$. Of course, this relation is based on independence. Using the eigenvalues of the covariance (which is symmetric and hence the eigenvectors are orthogonal), the expression could nevertheless provide some estimation of the upper bound of the initial condition on the boundary.
Of course, this is not necessary to quantify directly for this manuscript, but the Authors could incorporate this in their future studies on this matter.

AUTHOR answer: We agree this is not necessary to quantify for the present purposes, but in any case thank you for pointing this out to us, it may very well become useful for future work on this topic.

@@

\item
The errors in the figures, as well as the tolerances used in the paper (for instance $TOL_{res}$) seem to be large to me. Perhaps, they are not large relatively. The Authors should put this into its context, I think.

AUTHOR answer: We chose not to normalize the errors and estimators used in the numerics such that the a posteriori modeling-remainder estimator, $\eta_{rem,i}^k$ (shown in figures 6 and 13) is calculated exactly as given by Theorem 6.3, eq. (6.12d). The reason for such large values is that $\eta_{rem,i}^k$ is scaled by $G^{-\frac{1}{2}}$, where $G$ is the largest eigenvalue of the diffusion tensor $g$, which is chosen as a rather small constant for both examples (i.e., $5e-6$ and $2e-6$, respectively). We have added a sentence in Section 7, third paragraph where we comment on this.

@@

\item
Please, say something about a comparison of computing times of the heuristic method and the multiscale method in Sectino 7.3.2.

AUTHOR answer: A new remark has been added on p. 23 (remark (7.1)) where we address computation times of the heuristic and multi-scale methods.

\item
Figures 11 and 12 seem blurry (or is it my eyes after having read the manuscript).

AUTHOR answer: These figures do not appear blurry to the author (who possess eagle eyed vision, of course). Source files have been double checked, everything should be in order.

@@

\begin{enumerate}
\item
The authors discuss and propose a model for the speciation processes leading to new species. To do so, they make use of a discretized version of the above equations. The discussions on the speciation processes seems to be highly related to the choice of the discrete system and in particular on the reconstruction operator. No connection with the original continuous system is clearly explained neither how it depends on the discretization.
Speciation is modeled in term of clustering in the trait space. Does it depend on the traits kept in the model?

AUTHOR answer:

\item
From a modelling view point, how do you choose the selection function $\alpha=\alpha(x,y,t)$? How is it related to the coefficients $A_{i,j}$ in (2.5a)? How the multi-scale parameters are coupled?

AUTHOR answer: $A_{i,j}$ is related to $\alpha$ through eq. (2.6c), i.e., the effect of species $i$ on species $j$ (as opposed to individuals with different trait coordinates effect each other, as is the case at the population level). From a modeling point of view, $\alpha$ defines the ecological interactions in the system, and measures potentially competitive, mutualistic, and trophic interactions, and may in general be either symmetric or unsymmetric. Hence, it may take many different functional forms. For the numerical examples in the paper (which are only meant to illustrate the multi-scale framework, not represent any real eco-evolutionary system) we chose in the first case a symmetric competitive ecology, and in the second case an unsymmetric trophic ecology. More details on $\alpha$ can be found in [2].

The strategy for coupling the species and population scales in the multi-scale framework is outlined in section 5 "Speciation", the the main components being the mappings from the species scale to the population scale (reconstruction operator) and vice-versa (compression operator).
\end{enumerate}

@@

\textbf{References:}\\

$[$1$]$ Nordbotten, J. M., Bokma, F., Hermansen, J. S., \& Stenseth, N. C. (2020). "The dynamics of trait variance in multi-species communities". \textit{Royal Society open science}, 7(8), 200321.

$[$2$]$ Nordbotten, J. M., \& Stenseth, N. C. (2016). "Asymmetric ecological conditions favor Red-Queen type of continued evolution over stasis". \textit{Proceedings of the National Academy of Sciences}, 113(7), 1847-1852.

### 2021-08-09 manuscript/ex_article.tex
.

% REQUIRED
\begin{abstract}
This paper concerns the modeling and numerical simulation of the process of speciation. In particular, given conditions for which one or more speciation events within an ecosystem occur, our aim is to develop the modeling and simulation tools necessary to handle such events. The model we employ can be considered as an extension of the classical Lotka-Volterra model, where in addition to the species abundance, the model also governs the evolution of the species mean traits and species trait covariances, and in this sense generalizes the purely ecological Lotka-Volterra model to an eco-evolutionary model. Although the model thus allows for evolving species, by construction, it is not able to handle speciation events due to a breakdown of the underlying assumptions on which the model is derived (i.e., unambiguous identification of distinct species in the fitness landscape). Thus, the fundamental problem to overcome regarding speciation is that the unit of species is too coarse to capture the fine-scale dynamics of a speciation event. Our strategy is twofold; monitoring the species in the system and detecting speciation in advance, and, splitting the diverging `parent'-species into new `child'-species. Since the problem at hand is related to that of separate scales, it is convenient to have access to a fine-scale description of the same eco-evolutionary system. For this, we employ a trait-specific population density model governing the dynamics of the abundance density as a function of evolutionary traits. At this scale there is no a priori definition of species, but both species and speciation may be defined a posteriori as e.g., local maxima and saddle points of the population density, respectively.  As long as the identification of distinct species remains non-ambiguous, both the fine-scale (population) and coarse-scale (species) level models describe the same system dynamics (i.e., evolution of the same ecosystem). In this sense, we regard both species and speciation as emergent phenomena, ultimately due to individuals interacting within a changing environment. In practice, our strategy amounts to coupling the species and population level models in a multi-scale framework. This includes well-posedness and a posteriori error estimates of the time discrete population density model. Several numerical tests validate our framework in practice.
\end{abstract}

@@

N_{i,j}^k := (\mu_{B_{i,\nu}^{k,j}}^0(\Dcal N^k_i),\mu_{B_{i,\nu}^{k,j}}^1(\Dcal N^k_i),\mu_{B_{i,\nu}^{k,j}}^2(\Dcal N^k_i)).

@@

Figure \ref{fig1:bi} below shows the splitting of an example density region in 2D, where $x_i^k = (0,0)$ and $\upsilon_i^k = \diag(\upsilon_{i,11}^k, \upsilon_{i,22}^k)$.

@@

s_{h,i}^k := \Dcal N^k_i 

@@

where $\Dcal : \real^{1+d+d^2} \rightarrow C^\infty(\Omega) \cap X_k$ is the density reconstruction operator (i.e., macro-to-micro scale mapping), such that $\Dcal N^k_i$ is any smooth distribution characterized by a mean and variance, and adhering to the relevant boundary condition (in this case, homogenous Dirichlet). Thus, the global ($s$-species) abundance density reconstruction is $H^1$-conforming  in space i.e., 

@@

We assume the discretization errors associated with solving the models are negligible compared to the modeling error which we are interested in, hence in practice we calculate only the remainder estimator $\eta_{\textn{rem},i}^k$ from \eqref{etarem} when estimating~\eqref{finalestimate}. Moreover, the density reconstruction operator $\Dcal$ is implemented as a normal distribution, i.e., for a species $N_i^k = (n_i^k,x_i^k,\upsilon_i^k)$ we have

>>>>

% REQUIRED
\begin{abstract}
\red{
This paper concerns the modeling and numerical simulation of the process of speciation. In particular, given conditions for which one or more speciation events within an ecosystem occur, our aim is to develop the necessary modeling and simulation tools. Care is also taken to establish a solid mathematical foundation on which our modeling framework is built. This is the subject of the first half of the paper. The second half is devoted to developing a multi-scale framework for eco-evolutionary modeling, where the relevant scales are that of species and individual/population, respectively. The species level model we employ can be considered as an extension of the classical Lotka-Volterra model, where in addition to the species abundance, the model also governs the evolution of the species mean traits and species trait covariances, and in this sense generalizes the purely ecological Lotka-Volterra model to an eco-evolutionary model. Although the model thus allows for evolving species, it does not (by construction) allow for the branching of species, i.e., speciation events. The reason for this is related to that of separate scales; the unit of species is too coarse to capture the fine-scale dynamics of a speciation event. Instead, the branching species should be regarded as a population of individuals moving along a selection of trait axes (i.e., trait-space). For this, we employ a trait-specific population density model governing the dynamics of the population density as a function of evolutionary traits. At this scale there is no a priori definition of species, but both species and speciation may be defined a posteriori as e.g., local maxima and saddle points of the population density, respectively.  
%As long as the identification of distinct species remains non-ambiguous, both the fine-scale (population) and coarse-scale (species) level models describe the same system dynamics (i.e., evolution of the same ecosystem), while  
%In this sense, we regard both species and speciation as emergent phenomena, ultimately due to individuals interacting within a changing environment. 
Hence, a system of interacting species can be described at the species level, while for branching species a population level description is necessary. 
%Assuming the existence of statistical moments of the population density function, and that the individuals of each species follows a known statistical distribution,  
The coupling of the species and population scales into a multi-scale framework is the subject of the second part of the paper. Moreover, since the population level model is formulated as a PDE, we first establish the well-posedness in the time-discrete setting, and then derive the a posteriori error estimates which provides a fully computable upper bound on an energy-type error, including also for the case of general smooth distributions. Our coupling strategy is twofold; monitoring the species in the system and detecting speciation in advance (using the a posteriori error bounds), and, splitting the diverging `parent'-species into new `child'-species. 
%Since the problem at hand is related to that of separate scales, it is convenient to have access to a fine-scale description of the same eco-evolutionary system. In practice, our strategy amounts to coupling the species and population level models in a multi-scale framework. This includes well-posedness and a posteriori error estimates of the time discrete population density model. 
Several numerical tests validate our framework in practice.
}
\end{abstract}

@@

N_{i,j}^k := (\mu_{B_{i,\nu}^{k,j}}^0(\Dcal N^k_i),\mu_{B_{i,\nu}^{k,j}}^1(\Dcal N^k_i),\mu_{B_{i,\nu}^{k,j}}^2(\Dcal N^k_i)),

@@

\red{where $\Dcal : \real^{1+d+d^2} \rightarrow C^\infty(\Omega) \cap X_k$ is the density reconstruction operator (i.e., macro-to-micro scale mapping), such that $\Dcal N^k_i$ is any smooth distribution characterized by a mean and variance, and adhering to the relevant boundary condition (in this case, homogenous Dirichlet).} Figure \ref{fig1:bi} below shows the splitting of an example density region in 2D, where $x_i^k = (0,0)$ and $\upsilon_i^k = \diag(\upsilon_{i,11}^k, \upsilon_{i,22}^k)$.

@@

s_{h,i}^k := \Dcal N^k_i. 

@@

%where $\Dcal : \real^{1+d+d^2} \rightarrow C^\infty(\Omega) \cap X_k$ is the density reconstruction operator (i.e., macro-to-micro scale mapping), such that $\Dcal N^k_i$ is any smooth distribution characterized by a mean and variance, and adhering to the relevant boundary condition (in this case, homogenous Dirichlet). 
Thus, the global ($s$-species) abundance density reconstruction is $H^1$-conforming  in space i.e., 

@@

We assume the discretization errors associated with solving the models are negligible compared to the modeling error which we are interested in, hence in practice we calculate only the remainder estimator $\eta_{\textn{rem},i}^k$ from \eqref{etarem} when estimating~\eqref{finalestimate} \red{(note that we do not normalize in the plots)}. Moreover, the density reconstruction operator $\Dcal$ is implemented as a normal distribution, i.e., for a species $N_i^k = (n_i^k,x_i^k,\upsilon_i^k)$ we have
